---
title: "SSE 708: Machine Learning in the Era of Big Data"
subtitle: "An introduction to Rstudio, tidyverse, and dplyr"
author: "David McCoy"
date: "July 7-11, 2025"
output: pdf_document
#  ioslides_presentation: default
#  slidy_presentation: default
editor_options: 
  chunk_output_type: console
---

### Learning objectives for today:

1. What is a data frame
2. How to read a comma separated values (CSV) file using `read_csv()` 
3. Get to know the data using `str()`, `head()`, `dim()`, and `names()` 
4. Manipulate the data frame using the R package `dplyr`'s main functions:
    - `rename()`
    - `select()`
    - `arrange()`
    - `filter()`
    - `mutate()`
    - `group_by()`
    - `summarize()`

### Readings

* Here are some  online resources (optional, but helpful!):
  * [Data Frames](https://learningstatisticswithr.com/lsr-0.6.pdf (See section 4.8))
  * [15 min intro to dplyr](https://www.youtube.com/watch?v=aywFompr1F4)
  * [Data wrangling cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)

### What is a data frame?

- A data frame is a data set. 
- We read data into R from common sources like Excel spreadsheets (.xls or .xlsx), 
    text files (.txt), comma separate value files (.csv), and other formats.
- The simplest format of data contains one row for each individual in the study. 
- The first column of the data identifies the **individual** (perhaps by a name or an ID **variable**). 
- Subsequent columns are **variables** that have been recorded or measured.

### Lake data from Baldi and Moore (B&M)

- Six rows of data from a study of mercury concentration across 53 lakes 
- I've added three fabricated rows 
- I've placed these data in Day-2 folder
- Let's find it there

### readr is a library to import data into R

- To access `readr`'s functions we load the library like this:

```{r}
library(readr)
```

- Click the green arrow to run the code or place your cursor on the line of code
and type `cmd + enter` (Mac) or `control + enter` (PC)
- A green rectangle that temporarily appears next to the code shows you that it 
has run.

### `read_csv()` to load the lake data in R
- `read_csv()` is a function from the `readr` library used to import csv files.
- code template: `your_data <- read_csv("pathway_to_data.csv")`
- The `<-` is called the **assignment operator**. It says to save the imported
data into an object called `your_data`.

```{r read-lake-data}
lake_data <- read_csv("Data_mercury_lake.csv")
```

* Anytime you see "##" on the html slides or in the PDF lecture files, the text in those
lines are the **output** of running the code in the previous line. So the lines 
above are the output displayed when you run the `read_csv()` function.

### Exercise 1

1. Execute the above code using either the green arrow or by clicking on it and 
hitting the keyboard shortcut (`cmd + enter` on mac or `Ctrl + enter` on PC). 
2. Note that the data appears in the Environment pane in the top right. 
   - Notice the number of **observations** and the number of **variables**. 
3. Click the tiny table icon to the right of the `lake_data` in the Environment 
pane to open the **Viewer** tab and inspect the data.

# Check your understanding!

### Four functions to get to know a dataset

* `head(your_data)`: Shows the first six rows of the supplied dataset
* `dim(your_data)`: Provides the number of rows by the number of columns
* `names(your_data)`: Lists the variable names of the columns in the dataset
* `str(your_data)`: Summarizes the above information and more

I use these functions all the time! Multiple times per session when working with
data to remind me what the variable names are, and what the data looks like.

### head()

First six rows:

```{r head-lake-data}
head(lake_data)
```

### dim()

```{r dim-lake-data}
dim(lake_data)
```

Are there 9 rows or columns of data?

### names()

```{r names-lake-data}
names(lake_data)
```

### str()

```{r str-lake-data}
str(lake_data)
```

### Using `dplyr` functions for data manipulation {.bigger} 

- `rename()`
- `select()`
- `arrange()`
- `filter()`
- `mutate()`
- `group_by()`
- `summarize()`

### Load the `dplyr` library to access the functions

```{r load-dplyr}
library(dplyr)
```

- These messages mean that some functions (e.g., filter()) share names with 
functions from other libraries. So, when we use `filter()` we will now use the
`dplyr` version because the `stats` library version has been masked.
- You don't need to worry about masking for now.

### Function 1: rename()

What do you think rename does?

First print the names of the variables:
```{r print-variable-names}
names(lake_data)
```

Run the `rename()` function and assign it to `lake_data_tidy`:
```{r}
lake_data_tidy <- rename(lake_data, name_of_lake = lakes)
```

Then reprint the variable names:
```{r}
names(lake_data_tidy)
```

### Function 1: rename() multiple variables at once

You can `rename()` multiple variables at once:

```{r rename-two-variables}
lake_data_tidy <- rename(lake_data, 
                         name_of_lake = lakes, 
                         ph_level = ph)
```

### Code template for rename() function

```{r rename, eval = F}
new_dataset <- rename(old_dataset, new_name = old_name)
```

Another way to write the above code is to use the **pipe** operator (`%>%`):

```{r rename-with pipe, eval = F}
new_dataset <- old_dataset %>% rename(new_name = old_name)
```

The pipe will become very useful in a few slides

### Function 2: select()

Based on the output below, what do you think `select()` does? 

```{r select-first-three}
smaller_data <- select(lake_data, lakes, ph, chlorophyll)
names(smaller_data)
```

### Function 2: select()

* We use `select()` to select a subset of **variables**. 
* This is very handy if we inherit a large dataset with several variables that
we do not need.

### Function 2: "negative select()"

We can also use "negative `select()`" to deselect variables. Suppose we wanted
to keep all variables except for `age_data`:

```{r deselect-last-variable}
smaller_data_2 <- select(lake_data, - age_data)
names(smaller_data_2)
```

We place a negative sign in front of `age_data` to remove it from the dataset.

### Rewrite using the pipe operator

```{r}
smaller_data <- lake_data %>% select(lakes, ph, chlorophyll)
smaller_data_2 <- lake_data %>% select(- age_data)
```

* Going forward, we will use the pipe operator to write code using any `dplyr` functions
* This is because we can use the pipe to stack many `dplyr` functions in a row

### Function 3: arrange()

What does arrange do? First type `View(lake_data)` to look at the original data. Then 
run the code and examine its output below. What is different?:

```{r arrange-ph}
lake_data %>% arrange(ph)
```

### Function 3: arrange() in descending order

```{r arrange-neg-ph}
lake_data %>% arrange(- ph)
```

### Function 3: arrange() by two variables

```{r arrange-age-and-ph}
lake_data %>% arrange(age_data, ph)
```

### Function 4: mutate()

- **`mutate()` is one of the most useful functions!** 
- It is used to add new variables to the dataset. Suppose that someone told you 
that the number of fish sampled was actually in hundreds, such that 5 is actually 500. You can use mutate to add a new variable to your dataset that is in the hundreds:

```{r mutate-fish-count}
lake_data_new_fish <- lake_data %>% 
  mutate(actual_fish_sampled = number_fish_sampled * 100)

lake_data_new_fish
```

### Use `%>%` to append several lines of code together
- We have saved many of new datasets in our environment! 
- If these datasets were larger, they would take up a lot of space. 
- Rather than saving a new dataset each time, we can make successive changes
to one dataset like this:

```{r rename-select-and-mutate}
tidy_lake_data <- lake_data %>% 
  rename(name_of_lake = lakes) %>%
  mutate(actual_fish_sampled = number_fish_sampled * 100) %>%
  select(- age_data, - number_fish_sampled)
```

- When you see `%>%`, say the words "**and then**...". For example, "Take `lake_data` **and then** rename `lakes` to name_of_lake, **and then** mutate..."

### Use `%>%` to "pipe" several lines of code together

```{r rename-select-and-mutate-eval}
tidy_lake_data <- lake_data %>% 
  rename(lake_name = lakes) %>%
  mutate(actual_fish_sampled = number_fish_sampled * 100) %>%
  select(- age_data, - number_fish_sampled)

tidy_lake_data
```

### Function 5: `filter()`

Filter is another very useful function! What might `filter()` do? 

### Function 5: `filter()`ing on numeric variables
We use filter to select which rows we want to keep in the dataset. Suppose you were only interested in lakes with `ph` levels of 7 or higher.

```{r filter-on-ph}
lake_data_filtered <- lake_data %>% filter(ph > 7)
lake_data_filtered
```

# Check your understanding!

### Function 5: `filter()`ing on character/string variables

Let's try a few more ways to `filter()` the data set since subsetting data is so 
important:

```{r filter-on-recent}
lake_data %>% filter(age_data == "recent")
```

- `==` is read as "is equal to"

```{r filter-not-on-recent}
lake_data %>% filter(age_data != "recent") 
```

- `!=` is read as "is not equal to"

### Function 5: `filter()`ing on character/string variables

```{r filter-on-name}
lake_data %>% filter(lakes %in% c("Alligator", "Blue Cypress")) 
```

- `%in%` is the "in" operator. We are selecting rows where the variable `lakes` 
belongs to the specified list.
- The `c()` combines "Alligator" and "Blue Cypress" into a list

### Function 5: multiple `filter()`s at once
```{r filter-multiple}
lake_data %>% filter(ph > 6, chlorophyll > 30)

#this is the same as:
lake_data %>% filter(ph > 6 & chlorophyll > 30)
```

- A comma or the "and" operator (`&`) are equivalent. Here they say, filter the dataset and keep only rows with `ph > 6` AND `chlorophyll > 30`

### Function 5: `filter()` using "or"
```{r filter-multiple-OR}
lake_data %>% filter(ph > 6 | chlorophyll > 30)
```

- `|` is the OR operator. At least one of `ph > 6` or `chlorophyll > 30` needs 
to be true.

### Functions 6 and 7: group_by() and summarize()

Let's execute the following code and see what it does.

```{r group_by-and-mean}
lake_data %>% 
  group_by(age_data) %>% 
  summarize(mean_ph = mean(ph))
```

What happened?

### Functions 6 and 7: group_by() and summarize()

Another one:

```{r group_by-and-mean-sd}
lake_data %>% 
  group_by(age_data) %>% 
  summarize(mean_ph = mean(ph),
            standard_deviation_ph = sd(ph))
```


### Exercise R1.1
Create a code block below that reads in the data for class (coursdata1_synth.csv), and ....

```{r}
set.seed(708)
n <- 120  # number of orders

sim_book_orders <- tibble(
  order_id     = sprintf("O%04d", 1:n),
  title        = sample(c("Data Science 101","R for Everyone","Causal Inference",
                          "Bayesian Basics","Deep Learning Hands-On","The Tidyverse",
                          "Statistical Thinking","Machine Learning Recipes"), n, TRUE),
  genre        = sample(c("CS","Statistics","AI","DataSci"), n, TRUE,
                        prob = c(.25,.25,.25,.25)),
  price_usd    = round(runif(n, 20, 80), 2),
  units        = sample(1:4, n, TRUE, prob = c(.5,.3,.15,.05)),
  member_status= sample(c("member","guest"), n, TRUE, prob = c(.6,.4)),
  country      = sample(c("USA","CAN","UK","AUS"), n, TRUE, prob = c(.55,.15,.2,.1)),
  rating       = sample(1:5, n, TRUE, prob = c(.05,.1,.2,.35,.3))
)

```

## Hands-on practice — dplyr & readr

We’ll use the **simulated bookstore orders** data you created earlier  
(`sim_book_orders.csv` in your working directory).

---

# Book Orders Data Analysis

## Setup
```{r setup, message=FALSE}
library(tidyverse)
library(forcats)
```

## 1. Import & Peek
Read "sim_book_orders.csv" into `books`. Run `head()`, `str()`, and `dim()`.

```{r import-data}
# Your code here

```

## 2. Rename Variables
Rename `price_usd` → `price`, `member_status` → `member`.

```{r rename-vars}
# Your code here

```

## 3. Create New Variables
Use `mutate()` to add:
- `revenue = price * units`
- `discount_flag = if_else(units > 2, "bulk", "regular")`

```{r create-vars}
# Your code here

```

## 4. Select a Tidy Column Order
Keep `order_id`, `title`, `genre`, `units`, `price`, `revenue`, `member`, `country`, `rating`.

```{r select-columns}
# Your code here

```

## 5. Filter Rows
Keep orders where `price > 50` and `genre == "AI"`.

```{r filter-data}
# Your code here

```

## 6. Arrange
Sort that filtered data by `revenue` (descending).

```{r arrange-data}
# Your code here

```

## 7. Group & Summarize
For each genre, compute average rating and total revenue.

```{r group-summarize-genre}
# Your code here

```

## 8. Member vs Guest Spend
Group by `member` and return the mean revenue and order count.

```{r member-analysis}
# Your code here

```

## 9. Re-order Factor Levels
Re-order `genre` by descending total revenue (hint: `fct_reorder()` from forcats).

```{r reorder-factors}
# Your code here

```

## 10. Export
Save the summary table from task 7 as "genre_revenue_rating.csv" using `write_csv()`.

```{r export-data}
# Your code here

```

### Recap: What functions did we use?

1. `library()` to load `readr` and `dplyr`.
2. `read_csv()` to read csv files from a directory.
3. `head()`, `str()`, `dim()`, and `names()` to look at our imported data.
4. `rename()` to rename variables in a data frame.
5. `select()` to select a subset of variables.
6. `arrange()` to sort a dataset according to one or more variables.
7. `mutate()` to create new variables.
8. `filter()` to select a subset of rows.
9. `group_by() and summarize()` to group the data by a categorial variable and 
calculate a statistic.
10. `mean()` and `sd()` to calculate the mean and standard deviation of variables.

### Recap: What operators did we use?

1. Assignment arrow (`<-`): This is our most important operator!
2. Greater than (`>`) There are also: 
    - Less than (`<`)
    - Greater than or equal to (`>=`), and,
    - Less than or equal to: (`<=`) 
4. Is equal to (`==`), and is not equal to (`!=`)
5. `%in%` to select from a list, where the list is created using `c()`, i.e., 
`lakes %in% c("Alligator", "Annie")`

