---
title: "SSE 708: Machine Learning in the Era of Big Data"
subtitle: "Substitution Estimators, TMLE"
author: "David McCoy"
date: "July 7-11, 2024"
output: pdf_document
editor_options:
  chunk_output_type: console
---

```{r setup,  include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Substitution Estimators

We start with simple substitution estimators for the average treatment affect (ATE).  The estimand is 
$E_W \{ E(Y \mid A=1,W)-E(Y \mid A=0,W) \}$,
which, under assumptions, is equal to the causal mean difference, $E(Y(1)-Y(0))$. 
In addition, we also estimate the causal attributable risk 
$EY-E_W \{E(Y \mid A=0,W)\}$ which is equal (under assumptions) to $E(Y-Y(0))$.  
Below are both simulations and data analysis.

```{r sub1}
### Data generating function
dat.gen.1=function(n,probsw,probsa,bY) {
  W=sample(1:5,n,replace=T,prob=probsw)
  A=rbinom(n,1,probsa[W])
  X=cbind(1,A,W-1)
  PYgivenAW=1/(1+exp(-(X%*%bY)))
  Y=rbinom(n,1,PYgivenAW)
  return(data.frame(W,A,Y))
}

### Parameters of Simulation
probsw=c(0.2,0.2,0.2,0.2,0.2)
probsa = c(seq(0.12,0.05,length=5))
bY = c(-2,1,-0.25)
n=1000
set.seed(403)
### Get simulated data
dat=dat.gen.1(n,probsw,probsa,bY)
### Display True Distribution
par(mfrow=c(1,1))
par(mar = c(5,5,2,5))
w=1:5
Xt=cbind(1,0,w-1)#A=0, everyone untreated
PYgivena0w=1/(1+exp(-(Xt%*%bY)))
Xt=cbind(1,1,w-1)
PYgivena1w=1/(1+exp(-(Xt%*%bY)))#A=1, everyone untreated
plot(w,PYgivena1w,pch="1",ylim=c(0.00,0.3),xlab="w",ylab="P(Y=1|A=a,W=w)")
points(w,PYgivena0w,pch="0")  
### Get matrix of relevant probabilities to calculate true ATE
yprobdat=data.frame(PYgivena0w,PYgivena1w)
yprobdat=data.frame(W=1:5,pw=probsw,yprobdat)
yprobdat=data.frame(yprobdat,probsa)

#### Get observed estimates in the simulated data Using Saturated (NP) Model
yprobobs=NULL
for(i in 1:5) {
  pyaw=mean(dat$Y[dat$A==1 & dat$W==i])
  pyaw=c(pyaw,mean(dat$Y[dat$A==0 & dat$W==i]))
  yprobobs=rbind(yprobobs,pyaw)
}
plot(w,yprobobs[,1],pch="1",ylim=c(0.00,0.4),xlab="w",ylab=expression(paste(hat(P),"(Y=1|A=a,W=w)",sep="")))
points(w,yprobobs[,2],pch="0")  

### Estimated Pn(W=w) - wtab just using simple proportions
tt=table(dat$A,dat$Y,dat$W)
wtab=(table(dat$W)/(length(dat$W)))
### tabout is estimated relevant probabilities for estimating ATE
tabout=NULL
for(i in 1:5) {
  tabout=rbind(tabout,c(wtab[i],tt[2,2,i],sum(tt[2,,i]),yprobobs[i,1],tt[1,2,i],sum(tt[1,,i]),yprobobs[i,2]))
}
colnames(tabout)=paste("est.",c("pw","ny1","na1","pya1w","ny0","na0","pya0w"),sep="")
############################################
## estimated data-generating distribution
############################################
#### Get all estimates and true values of data generating distribuiton in same data frame
res.out=cbind(yprobdat[,c(1:2,4,3)],diff.tr=yprobdat[,4]-yprobdat[,3],
              tabout,diff.est=tabout[,"est.pya1w"]-tabout[,"est.pya0w"])
#### Get estimates and true values of relevant parameters
##########  E(Y1),E(Y0),E(Y1-Y0),E(Y-Y0)
ATE.sat=sum(tabout[,1]*(tabout[,4]-tabout[,7]))
ey1.sat=sum(tabout[,1]*tabout[,4])
ey0.sat=sum(tabout[,1]*tabout[,7])
ATE.true=sum(yprobdat[,2]*(yprobdat[,4]-yprobdat[,3]))
ey0.true=sum(yprobdat[,2]*(yprobdat[,3]))
ey1.true=sum(yprobdat[,2]*(yprobdat[,4]))
ey.true =sum(yprobdat[,2]*yprobdat[,4]*yprobdat[,5]+(yprobdat[,2]*yprobdat[,3]*(1-yprobdat[,5])))
pim.true=ey.true-ey0.true #causal attributable risk
pim.sat=mean(dat[,"Y"])-ey0.sat #observed attributable risk
### Print out values
####################  True
## E(Y(0))
print(ey0.true)
## E(Y(1))
print(ey1.true)
## EY
print(ey.true)
## ATE
print(ATE.true)
## PIM
print(pim.true)
####################  Estimated
## EY(0)
print(ey0.sat)
## EY(1)
print(ey1.sat)
## EY
print(mean(dat[,"Y"]))
## ATE
print(ATE.sat)
## ATE
print(pim.sat)
#######################  Repeat assuming parametric model for Q(A,W)=E(Y|A,W)
#########
########  Note, not saturated model
glm1=glm(Y~A+factor(W),data=dat,family=binomial())
summary(glm1)
############################################
## Figure for estimated data-generating distribution
############################################
datn=data.frame(W=1:5,A=0)
predA0=predict(glm1,newdata=datn,type="response")
datn=data.frame(W=1:5,A=1)
predA1=predict(glm1,newdata=datn,type="response")
par(mfrow=c(1,1))
plot(w,yprobobs[,1],pch="1",ylim=c(0.00,0.50),xlab="w",ylab=expression(paste(hat(P),"(Y=1|A=a,W=w)",sep="")))
points(w,yprobobs[,2],pch="0")
lines(w,predA0,lty=1,col=1)
lines(w,predA1,lty=2,col=2)

### Add this to results data frame
res.out=cbind(res.out,glm.predA1=predA1,glm.predA0=predA0,glm.diff=predA1-predA0)

### Get substitution estimate
datn=dat
predA=predict(glm1,newdata=datn,type="response")
eyA.glm=mean(predA)#used for calculating attribute risk
datn[,"A"]=0
predA0=predict(glm1,newdata=datn,type="response")
eyA0.glm=mean(predA0)
datn[,"A"]=1
predA1=predict(glm1,newdata=datn,type="response")
eyA1.glm=mean(predA1)
ATE.glm=eyA1.glm-eyA0.glm
pim.glm=eyA.glm-eyA0.glm
# EY(0)
print(eyA0.glm)
# EY(1)
print(eyA1.glm)
# EY
print(eyA.glm)
# ATE
print(ATE.glm)
# PIM
print(pim.glm)
#### Combine all results into the final results data frame
xx=c(NA,NA,ey1.true,ey0.true,ATE.true,NA,NA,NA,ey1.sat,NA,NA,ey0.sat,ATE.sat,eyA1.glm,eyA0.glm,ATE.glm)
res.out2=rbind(res.out,xx)
res.out2=data.frame(res.out2)
res.out2[,1]=as.character(res.out2[,1])
res.out2[6,1]="Marginals"
```

Now, we examine the WCGS data


```{r sub2}
## Package that allows one to read in data from STATA and other stat software
library(foreign)
## Make sure that wcgs.dta is in the right directory
wcgs=read.dta("wcgs.dta")
### reduce to only variables needed for analysis and also get rid of rows with missing
### values
dat=na.omit(wcgs[,c(11,2:7,9,10)])
##############  ATE for binary behavior type:  dibpat0=1 (type A), dibpat0=0 (type B).
#############  Assume for now, remaining variables are confounders
glm.all=glm(chd69~.,data=dat,family=binomial())
X=dat[,-1]
### Get estimate E(Y|A=0,W)
X[,"dibpat0"]=0
predY.A0=predict(glm.all,type="response",newdata=X)
### Get estimate E(Y|A=1,W)
X[,"dibpat0"]=1
predY.A1=predict(glm.all,type="response",newdata=X)
#### estimates of EY0,EY1 and E(Y1-Y0)
est.EY0=mean(predY.A0)
est.EY1=mean(predY.A1)
est.ATE=est.EY1-est.EY0
print(c(est.EY0,est.EY1,est.ATE))
###################   Repeat using SL as predictor
#### Need to have one vector of outocme, and a matrix of predictors
Y=dat[,1]
X=dat[,-1]
library(SuperLearner)
#### Add a learner
SL.inter2=function (Y, X, newX, family, direction = "both", trace = 0, 
                    k = 2, ...) 
{
  fit.glm <- glm(Y ~ ., data = X, family = family)
###########  Make the null model at least include the variable of interest (dibpat0)
  fit.step <- step(fit.glm, scope = list(lower=Y~dibpat0,upper=Y ~ .^2), direction = direction, 
                   trace = trace, k = k)
  pred <- predict(fit.step, newdata = newX, type = "response")
  fit <- list(object = fit.step)
  out <- list(pred = pred, fit = fit)
  class(out$fit) <- c("SL.step")
  return(out)
}

SL.library <- c("SL.glm", "SL.stepAIC", "SL.glmnet", "SL.gam", "SL.mean","SL.knn","SL.inter2")
SL.wcgs=SuperLearner(Y=Y,X=X,family=binomial(),SL.library=SL.library,verbose=F)
SL.wcgs
### Get estimate E(Y|A=0,W)
Xnew=X
Xnew[,"dibpat0"]=0
predY.A0=predict(SL.wcgs,newdata=Xnew,X=X,Y=Y)
### Get estimate E(Y|A=1,W)
Xnew[,"dibpat0"]=1
predY.A1=predict(SL.wcgs,newdata=Xnew,X=X,Y=Y)
#### estimates of EY0,EY1 and E(Y1-Y0)
est.EY0.SL=mean(predY.A0$pred)
est.EY1.SL=mean(predY.A1$pred)
est.ATE.SL=est.EY1.SL-est.EY0.SL
out.wcgs=cbind(c(est.EY0,est.EY1,est.ATE),c(est.EY0.SL,est.EY1.SL,est.ATE.SL))
rownames(out.wcgs)=c("EY0","EY1","ATE")
colnames(out.wcgs)=c("glm","SL")
print(out.wcgs)
```


# TMLE
Let's practice on WCGS data

```{r tmle2}
library(foreign)
library(tmle)
#setwd("~/Dropbox (UC Berkeley Biostat)/hubbardlap/Michigan Summer 2016/Class Material/data/")
wcgs=read.dta("wcgs.dta")
### reduce to only variables needed for analysis
dat=wcgs[,c(11,2:7,9,10)]
summary(dat)
Y=dat[,1]
A=dat[,9]
W=dat[,-c(1,9)]
### Note, only missing data for chol0 (adjustment variable).  Use trick
### to make new basis functions for cholestoral 
### (say Z) =(I(Z is obs),I(Z is obs)*Z).
### Below is general code one can use on the so-called W matrix
xp=dim(W)[2]
sum.na=function(x){sum(is.na(x))}
sna=apply(W,2,sum.na)
nmesW=names(W)
for(k in 1:xp){
    if(sna[k] > 0) {
      ix=as.numeric(is.na(W[,k])==F)
    	W[is.na(W[,k]),k]=0 #replace the default NA with 0
    	W=data.frame(W,ix)
    	nmesW=c(nmesW,paste("Imiss_",nmesW[k],sep="")) }
} 
names(W)=nmesW
W[W$chol0==0,]
table(is.na(W$chol0))

#### Library for SL 
# Below is a good library, but just use simpler one so it runs faster - this library can
## take several minutes to complete.
#Q.lib= c("SL.glm", "SL.stepAIC", "SL.glmnet", "SL.gam", "SL.glm.interaction",
#         "SL.earth", "tmle.SL.dbarts2")
Q.lib= c("SL.glm", "SL.glmnet", "SL.gam")
## Below can take a bit even with smaller, not very good, library
tmle.wcgs=tmle(Y,A,W,family="binomial",Q.SL.library=Q.lib, verbose = T)
print(tmle.wcgs)
### Get relevant estimates and CI
EY0.tmle=mean(tmle.wcgs$Qstar[,1])
EY1.tmle=mean(tmle.wcgs$Qstar[,2])
ATE.tmle=tmle.wcgs$estimates$ATE$psi
CI.tmle=tmle.wcgs$estimates$ATE$CI
### Compare estimates
out.wcgs=data.frame(out.wcgs,TMLE=c(EY0.tmle,EY1.tmle,ATE.tmle))
 ########### Print results.
out.wcgs
```

